Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-18 manuscript_agu_rev1.tex
Move rs info to methods

\subsection{Voyage and Station Data}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. Derived thermodynamic [virtual potential temperature ($theta_v$), RH, lifting condensation level, etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the lifting condensation level (LCL) distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models can cause more boundary mixing across heights and thus lower near-surface RH.

@@

In order to examine the potential link in the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S available from the 24 RV \emph{Polarstern} voyages and the MARCUS, NBP1704, TAN1702, and TAN1802 campaigns. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections~\ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). We focus on comparing $\theta_v$ and RH, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of $\theta_v$ and RH are shown in Fig.~\ref{fig:potential-temperature}.

>>>>

\subsection{Voyage and Station Data}
\label{sec:voyage-and-station-data}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. In total, about 2300 radiosonde profiles south of 40°S were available. Spatially and temporally colocated profiles were taken from the models. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. Derived thermodynamic [virtual potential temperature ($\theta_v$), lifting condensation level (LCL), etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the LCL distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models can cause more boundary mixing across heights and thus lower near-surface RH.

@@

In order to examine the potential link in the cloud biases to the local physical conditions, we analyzed the radiosonde profiles available from the campaigns (Section~\ref{sec:voyage-and-station-data}). The profiles were partitioned into the same subsets as above (Sections~\ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). We focus on comparing $\theta_v$ and RH, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of $\theta_v$ and RH are shown in Fig.~\ref{fig:potential-temperature}.
